---
title: "Code for the paper Megaliths of Monte Penide and environs"
author: "Miguel Carrero, Andrew Bevan, Mark Lake"
date: "8 February 2019"
output: html_document
---

# Setup

```{r echo=FALSE, results='hide'}
numSims <- 4

spatpack<-c("raster","spatstat","rgdal","maptools")
lapply(spatpack, require, character.only=TRUE)
```

The results are based on `r numSims` simulations.

The following libraries were loaded: `r spatpack`.


# PART 3 Data and methods

## Load data
```{r results='hide'} 
elevation <- raster("grids/dem.tif")
eucl_dist_wat <- raster("grids/wsheddists.tif")
studyarea <- readOGR(dsn="shp/studyarea", layer="studyarea")
sites <- read.table(file="csv/sites.csv",header=TRUE, sep=";")
area <- as(studyarea,"owin")
sppp <- ppp(x=sites$x, y=sites$y, window=area)

```

# Part 4 First-Order Location Model

## Summarise univariate relationships 

```{r results='hide'} 
elev <- as.im(as(elevation,"SpatialGridDataFrame"))
watersheddist <- as.im(as(eucl_dist_wat,"SpatialGridDataFrame"))
elev.rh <- rhohat(sppp, elev, confidence=0.95)
water.rh <- rhohat(sppp, watersheddist, confidence=0.95)

```
```{r}
pdf(width=9, height=7, paper="a4r", file="output/fig3cd.pdf")
par(mfrow=c(1,2))
par(mar=c(3, 3, 1, 1)) #c(bottom, left, top, right)
plot(elev.rh, main="", xlab="metres", ylab="", ylim=c(0,2.2e-06), legend=FALSE, cex.axis=0.7)
legend("topleft", legend="Elevation", cex=0.7, bty='n', text.font=2)
plot(water.rh, main="", xlab="metres", ylab="", ylim=c(0,2.2e-06), legend=FALSE, cex.axis=0.7)
legend("topright", legend="Watershed Distance", cex=0.7, bty='n', text.font=2)
par(mfrow=c(1,1))
dev.off()
```

The plot is saved as figs 3c and 3d in the folder 'output'.

## Calculate the first order logistic model

```{r} 
covlist <- list(elev, watersheddist)
names(covlist) <- c("elev", "watersheddist")
fotrend <- ~ elev + watersheddist
mod1 <- step(ppm(sppp, trend=fotrend, interaction=NULL, covariates=covlist, method="logi"))
summary(mod1)
```

```{r}
pdf(width=9, height=7, paper="a4r", file="output/fig3e.pdf")
logodds <- -16.379939413+(elev*0.006976940)+(watersheddist*-0.001089358)
par(mfrow=c(1,1))
plot(logodds, main="")
dev.off()
```

The plotted model is saved as fig3e in the folder 'output'.

# PART 5 Second order clustering

## Random (null) model

Calculate the inhomogeneous pcfunction with a 95% envelope from a
wholly random Poisson process, *mod0*.

```{r results='hide'}
mod0 <- ppm(sppp, ~1) #Null model
Pcfinhom_mod0 <- envelope(mod0, fun=pcfinhom, correction="iso", nsim=numSims)
```
```{r echo=FALSE}
pdf(width=9, height=7, paper="a4r", file="output/fig4a.pdf")
plot(Pcfinhom_mod0, xlim=c(0,8000),ylim=c(0, 30), legend=FALSE, main="a. Random model")
dev.off()
```

The plot is saved in the folder 'output' as fig4a.

## First order model

Calculate the inhomogeneous pcfunction with a 95% envelope conditioned
by the first order logistic model *mod1*.

```{r results='hide'}
Pcfinhom_mod1 <- envelope(mod1, fun=pcfinhom, correction="iso", nsim=numSims)
```

```{r}
pdf(width=9, height=7, paper="a4r", file="output/fig4b.pdf")
plot(Pcfinhom_mod1, xlim=c(0,8000),ylim=c(0, 30), legend=FALSE, main="b. First-order model")
dev.off()
```

The plot is saved in the folder 'output' as fig4b.

## First and second order (area interaction) model

Calculate the inhomogeneous pcfunction with a 95% envelope conditioned
by a point process model, *mod2*, which combines the first order
logistic model and a second order area interaction model.

Note that it can take a long time to run this code with a large number
of simulations.

```{r results='hide'}
mod2 <- step(ppm(sppp, trend=fotrend, interaction=AreaInter(1500), covariates=covlist, method="logi"))
Pcfinhom_mod2 <- envelope(mod2, fun=pcfinhom, correction="iso", nsim=numSims)
```

```{r}
pdf(width=9, height=7, paper="a4r", file="output/fig4c.pdf")
plot(Pcfinhom_mod2, xlim=c(0,8000),ylim=c(0, 30),legend=FALSE, main="c. First and second-order model")
dev.off()
```
   
The plot is saved in the folder 'output' as fig4c.
   
## Compare models

We can check the AIC's for the models.

	Model                          | AIC
	-------------------------------|--------------
	Null (mod0)                    | `r AIC(mod0)`
	First-order (mod1)             | `r AIC(mod1)`
	First and second order (mod2)  | `r AIC(mod2)`


# PART 6 Mound size and shape

## Correlation analysis between site distance and volume

Calculate the mark correlation function to explore the spatial
clustering of mounds by volume. 

```{r results='hide', warning=FALSE}

area <- as(studyarea,"owin")
points <- ppp(x=sites$x,y=sites$y, marks=sites$volume, window= area)
mcorr1 <- envelope (points, markcorr, nsim=numSims)
pdf(width=9, height=7, paper="a4r", file="output/fig5.pdf")
plot(mcorr1, main="mark correlation function", legend=FALSE)
dev.off()
```

The plot is saved in the folder 'output' as fig5.

## Hierarchical distribution of mound sizes

Here we calculate the mean and/sum of ranks in each level if we
hierarchically allocate tombs to each level.  In other words, the
biggest tomb in each of groups 1-9 is placed in level 1, then the
second biggest tomb in each of groups 1 to 9 is placed in level 2, and
so on.  We can then compare mean and/or sum of ranks in each level
with what we would get if we simply allocate tombs to levels by rank
with no reference to their group.  Since not all groups are equally
represented, we adjust the expectation accordingly, i.e. the 'levels'
will not have equal numbers of members.

### Functions for analysis

#### A function to calculate the ranks for each group.

```{r}
partition.ranks.by.group <- function(group.ids, tombs, shuffle=FALSE) {
    group.ranks <- list()
    if (shuffle) {
        rankvol <- sample(tombs$rankvol)
    }
    else
    {
        rankvol <- tombs$rankvol
    }    
    for (g in group.ids) {
        group.ranks[[g+1]] <- rankvol[which(tombs$group %in% c(g))]
    }
    return(group.ranks)
}
```

#### A function to allocate ranks to levels according to group membership
		
This iterates over groups allocating ranks to each level.  Once a
group is exhausted (i.e. there are no more tombs) we simply skip that
group.  In this case we store the ranks in one list per level because
we are actually interested in aggregate properties of ranks per level,
i.e. we regard [1, 2, 3][4, 5, 6] and [2, 1, 3][6, 5, 4] as matches in
the sense that they would have the same sum and/or mean.

```{r}
allocate.ranks.to.levels.by.group <- function (groups, group.ranks, max.tombs.per.group) {
    ranks.in.levels <- list()
    level <- 1
    while (level <=  max.tombs.per.group) {
        tmp.ranks <- rep(NA, length(groups$gis))
        i <-  1
        for (g in groups$gid) {
            if (groups$ntombs[g+1] >= level) {
                                        # So there is at least one
                                        # tomb left in this group
                tmp.ranks[i] <- group.ranks[[g+1]][level]
                i <-  i + 1
            }
        }
        ranks.in.levels[[level]] <- tmp.ranks
        level <- level+1
    }
    return (ranks.in.levels)
}
```	

#### A function to allocate ranks *ignoring* group membership

```{r}
allocate.ranks.to.levels.ignoring.group <- function(ranks.per.level.by.group, tombs, max.tombs.per.group) {
    ranks.in.levels <- list()
    level <- 1
    total.allocated <- 0
    global.rank <- 0
    while (level <= max.tombs.per.group) {
        tmp.ranks <- rep(NA, ngroups)
                                        # How many to allocate at this level
                                        # to match expected
        n.to.allocate <- length (ranks.per.level.by.group[[level]][which (! is.na(ranks.per.level.by.group[[level]]))])
        total.allocated <- total.allocated + n.to.allocate
        i <- 1
        while (i <= n.to.allocate) {
            global.rank <- global.rank + 1
            tmp.ranks[i] <- tombs$rank[global.rank]
            i <- i + 1
        }
        ranks.in.levels[[level]] <- tmp.ranks
        level <- level+1
    }
    return (ranks.in.levels)
}
```

#### Functions to compute the sum and mean rank at each level

```{r}
sum.ranks.by.level <- function (ranks.by.level, nlevels) {
    ranks.sums <- rep(1, nlevels, NA)
    level <- 1
    for (level in seq(1, nlevels, 1)) {
        ranks.sums[level] <- sum (ranks.by.level[[level]], na.rm=T)
    }
    return (ranks.sums)
}

mean.ranks.by.level <- function (ranks.by.level, nlevels) {
    ranks.means <- rep(1, nlevels, NA)
    level <- 1
    for (level in seq(1, nlevels, 1)) {
        ranks.means[level] <- mean (ranks.by.level[[level]], na.rm=T)
    }
    return (ranks.means)
}
```

### Load the data and get basic information

Load the data and drop tombs with volume = zero

```{r results='hide'}
tombs <- sites[which(sites$volume!=0),]
```

Obtain information about the number of groups and the number of
tombs per group.

```{r results='hide'}
ntombs <- length(tombs$id)

groups.id <- sort(unique(tombs$group))
ngroups <- length(groups.id)
groups.hist <- hist(tombs$group, breaks=seq(min(groups.id)-0.5, max(groups.id)+0.5, 1), plot=FALSE)

groups <- data.frame("gid"=groups.id, "ntombs"=groups.hist$count)
groups$min.vol <- sapply(groups.id, function(x) min(tombs$volume[which(tombs$group %in% c(x))]))
groups$mean.vol <- sapply(groups.id, function(x) mean(tombs$volume[which(tombs$group %in% c(x))]))
groups$med.vol <- sapply(groups.id, function(x) median(tombs$volume[which(tombs$group %in% c(x))]))
groups$max.vol <- sapply(groups.id, function(x) max(tombs$volume[which(tombs$group %in% c(x))]))
max.tombs.per.group <- max(groups$ntombs)
```

There are `r ntombs` tombs distributed acrosss  `r ngroups` groups, which
have the following characteristics in terms of tomb numbers and volumes:

```{r echo=FALSE}
groups
```

We see that groups 1, 3, 4, 5, 6, and 8 have means of similar order of
magnitude.  Groups 0, 2 and 7 are outliers, but also have fewer than 4
tombs.  There may be grounds to omit 0, 2 and 7 from analysis of
hierarchy, but we have not done so.


### Do the analysis

First, prepare the volume ranks (in descending order) and find the ranks in each group.

```{r results='hide'}
tombs$rankvol <- rank(-tombs$volume)
ranks.by.group <- partition.ranks.by.group(groups$gid, tombs, FALSE)
```

Now consider the observed situation.  We allocate the top rank in each
group to level 1, then the second rank in each group to level 2, and
so on.

```{r results='hide'}
observed.ranks.per.level <- allocate.ranks.to.levels.by.group (groups, ranks.by.group, max.tombs.per.group)
```

Next, consider the ideal situation.  In other words, we allocate the
top $N_1$ ranks to level 1, where $N$ is the number of groups which
contribute to that level.  We then allocate the next $N$ ranks to
level 2, where $N_2$ is the number of groups which contribute to that
level, and so on.

```{r results='hide'}
ideal.ranks.per.level <- allocate.ranks.to.levels.ignoring.group (observed.ranks.per.level, tombs, max.tombs.per.group) 
```

Now collate some summary information.  Specifically, the sum of ranks
and mean rank at each level under the two scenarios.

```{r results='hide'}
resultsB <- data.frame("level"=seq(1, max.tombs.per.group, 1))
resultsB$observed.sums <- sum.ranks.by.level (observed.ranks.per.level, max.tombs.per.group)
resultsB$observed.means <- mean.ranks.by.level (observed.ranks.per.level, max.tombs.per.group)
resultsB$ideal.sums <- sum.ranks.by.level (ideal.ranks.per.level, max.tombs.per.group)
resultsB$ideal.means <- mean.ranks.by.level (ideal.ranks.per.level, max.tombs.per.group)
```

```{r echo=FALSE}
resultsB
```

What we see here is that the ideal mean ranks increase from one level
to the next whereas that is not always true of the observed means.  We
also see that the level 1 and 2 ideal mean ranks are substantially
smaller than their observed equivalents.

We learn from this that the allocation of tomb volumes across groups
is not *perfectly* hierarchical.  However, it is still worth
considering whether the observed distribution of mean ranks per level
is closer to the ideal distribution than might be expected by chance.
We can examine this by re-labelling (randomly shuffling the tomb
volumes) and re-running the allocation of ranks to levels by group.

```{r results='hide'}
resultsB.sim.sums <- data.frame("level"=seq(1, max.tombs.per.group, 1))
resultsB.sim.means <- data.frame("level"=seq(1, max.tombs.per.group, 1))
nsims <- 999
for (sim in seq(1, nsims, 1)) {
    sim.ranks.by.group <- partition.ranks.by.group(groups$gid, tombs, TRUE)
    sim.ranks.per.level <- allocate.ranks.to.levels.by.group (groups, sim.ranks.by.group, max.tombs.per.group)
    command <- paste ("resultsB.sim.sums$s", sim, " <- sum.ranks.by.level(sim.ranks.per.level, max.tombs.per.group)", sep="")
    eval(parse(text=command))
    command <- paste ("resultsB.sim.means$s", sim, " <- mean.ranks.by.level (sim.ranks.per.level, max.tombs.per.group)", sep="")
    eval(parse(text=command))
}
```

We can now find out how the observed mean ranks at each level compare
to the simulated mean ranks at each level.  One way to do this is to
compute the rank of the observed mean ranks among the simulated mean
ranks.

```{r results='hide'}
nlevels <- length(resultsB.sim.means$s1)
tmp.rank.observed <- rep(NA, nlevels)
tmp.min.rank.sims <- rep(NA, nlevels)
tmp.pval <- rep(NA, nlevels)
for (level in seq(1, nlevels, 1)) {
    tmp.rank.observed[level] <- rank(c(resultsB$observed.means[level], as.numeric(resultsB.sim.means[level, 2:(nsims+1)])))[1]
    tmp.pval[level] <- tmp.rank.observed[level] / (nsims + 1)
    tmp.min.rank.sims[level] <- min(as.numeric(resultsB.sim.means[level, 2:(nsims+1)]))
}
resultsB$observed.rank.among.sims <- tmp.rank.observed
resultsB$pval <- tmp.pval
resultsB$min.rank.among.sims <- tmp.min.rank.sims
rm(tmp.rank.observed, tmp.min.rank.sims, tmp.pval)
```

```{r echo=FALSE}
resultsB[,c(1,6,7)]
```

This is an interesting result.  The first two hierarchical levels are
significant at $p <= 0.05$, and 3 out of the the first 4 are significant
at $p <= 0.05$.  The largest tombs are distributed across the groups in
a way that is broadly (even if not perfectly) hierarchical to an
extent that is unlikely to occur by chance alone.

